Hyperbolic SMC: a Variational Proposal for Sequential Monte Carlo Phylogenetic Inference

### 2024-07-29 paper/paper.tex
Fix typo
Add more content

% additional packages
\usepackage{siunitx}
\usepackage{physics}

@@

@@

\begin{enumerate}
  \item How are biological sequences (tree tips) embedded into latent space?
  \item Given the embeddings of the $N$ tree tips, and prescribed a tree topology, how are the $N-1$ internal nodes embedded?
  \item For a given branch, how to the embeddings of the two terminal nodes determine the distribution from which the branch length is sampled?
\end{enumerate}

@@

@@

>>>>

% additional packages
\usepackage{siunitx}
\usepackage{physics}
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\subsection{Motivation}

@@

\begin{enumerate}
  \item How are biological sequences (tree tips) embedded into latent space?
  \item Given the embeddings of the $N$ tree tips, and prescribed a tree topology, how are the $N-1$ internal nodes embedded?
  \item For a given branch, how do the embeddings of the two terminal nodes determine the distribution from which the branch length is sampled?
\end{enumerate}

@@

At each coalescent event, two trees from the forest are selected uniformly to be merged under a new internal node. Hyperbolic SMC requires a deterministic function that maps the embeddings of two tree roots to the embedding of the new internal node. A reasonable choice is the midpoint of the geodesic between the two roots. However, a desirable property is that the internal node falls closer to the origin than either of the two roots. We want trees to grow outward so that child nodes can leverage the increased space toward the boundary of the unit ball.

% TODO run an experiment to test different internal embedding functions
An improved heuristic is to first project the further root node onto the same hypersphere as the closer root node. Then, the geodesic midpoint is taken as the new internal node. A key requirement of the embedding function is that internal node embeddings are differentiable with respect to the embeddings of their subtree roots. This allows all internal embeddings to be optimized indirectly via gradient steps on the tip embeddings.

% TODO write up the math for the geodesic midpoint, which is simpler in the case that the two roots are already on the same hypersphere

@@

% TODO test log normal vs "squared normal" again
Ultimately, after selecting two trees to merge at a coalescent event, we must manifest the two branch length distributions. Whichever class of distributions is used, the child-parent embedding distances should govern the mean of the distribution. The original VCSMC proposal uses exponential distribution. However, we adopt the commonly used log-normal distribution.

At each coalescent event, the two branch lengths are sampled independently from log-normal distributions with means equal to the embedding distances and variances equal to a global learnable parameter. The reparameterization trick is used to propagate gradients through the sampling process to the embeddings.

\subsection{Implementation}

Algorithm \ref{alg:hyperbolic-smc} outlines how the Hyperbolic SMC proposal fits into the VCSMC framework.

% TODO replace this with Hyperbolic SMC
\begin{algorithm}
  \caption{Hyperbolic Sequential Monte Carlo}
  \label{alg:hyperbolic-smc}
  \begin{algorithmic}[1]
    \Require $\mathbf{Y} = \{Y_1, \cdots, Y_M\} \in \Omega^{N \times M}$, $\theta = (Q, \{\lambda_i\}_{i=1}^{|E|})$
    \State \textbf{Initialization.} $\forall k, s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
    \For{$r = 1$ to $R = N - 1$}
    \For{$k = 1$ to $K$}
    \State \textbf{Resample} $\mathbb{P}(a_{r-1}^k = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l}$
    \For{$i = 1$ to $L = \binom{N-r}{2}$}
    \For{$m = 1$ to $M$}
    \State \textbf{Form look-ahead partial state}
    \[
      s_r^{k,m}[i] \sim q(\cdot|s_{r-1}^{a_{r-1}^k})
    \]
    \State \textbf{Compute potentials}
    \[
      w_r^{k,m}[i] = \frac{\pi(s_r^{k,m}[i]) \cdot \nu^{-}(s_{r-1}^{a_{r-1}^k})}{\pi(s_{r-1}^{a_{r-1}^k}) q(s_r^{k,m}[i] | s_{r-1}^{a_{r-1}^k})}
    \]
    \EndFor
    \EndFor
    \State \textbf{Extend partial state}
    \[
      s_r^k = s_r^{k,J}[I], \quad \mathbb{P}(I = i, J = j) = \frac{w_r^{k,j}[i]}{\sum_{l=1}^L \sum_{m=1}^M w_r^{k,m}[l]}
    \]
    \State \textbf{Compute weights}
    \[
      w_r^k = \frac{1}{ML} \sum_{i=1}^L \sum_{m=1}^M w_r^{k,m}[i]
    \]
    \EndFor
    \EndFor
    \Ensure $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
  \end{algorithmic}
\end{algorithm}
